---
title: "Final Project Step 3: Ensemble Tree Methods for House Price Prediction"
author: "Group 6: Cristhian Cruz, Julian McKinley, Luisa Ribeiro, Matheus Lucena"
date: "2026-04-13"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  results = "hold"
)
```

# Problem Statement
The goal of this project is to predict house sale price using housing characteristics from the King County housing dataset. This is a regression problem, since the response variable is continuous.
This problem matters because accurate house price prediction is useful in real estate valuation, pricing strategy, and market analysis. In this final step of the project, we focus on ensemble tree methods from Chapter 8 
and compare:

Bagging
Random Forest
Boosting

A single regression tree is also included as a benchmark, but the main emphasis is on the three required ensemble methods.

# Packages

```{r}
suppressPackageStartupMessages({
  library(tidyverse)
  library(caret)
  library(rpart)
  library(rpart.plot)
  library(randomForest)
  library(gbm)
  library(knitr)
})
```

# Helper Functions
```{r}
set.seed(5703)

price_metrics <- function(pred_log, actual_price) {
  pred_price <- exp(as.numeric(pred_log))
  mse <- mean((actual_price - pred_price)^2)
  rmse <- sqrt(mse)
  mae <- mean(abs(actual_price - pred_price))
  r2 <- 1 - sum((actual_price - pred_price)^2) /
    sum((actual_price - mean(actual_price))^2)
  
  tibble(
    RMSE = rmse,
    MAE = mae,
    MSE = mse,
    R2 = r2
  )
}

print_price_metrics <- function(model_name, pred_log, actual_price) {
  out <- price_metrics(pred_log, actual_price)
  cat(model_name, "\n")
  cat("RMSE:", round(out$RMSE, 3), "\n")
  cat("MAE :", round(out$MAE, 3), "\n")
  cat("MSE :", round(out$MSE, 3), "\n")
  cat("R^2 :", round(out$R2, 6), "\n\n")
  invisible(out)
}
```

# Data Description and Loading
The dataset used is kc_house_data.csv. The response variable is price. Because house prices are strongly right-skewed, we model log(price) and then transform predictions back to the original price scale for evaluation.
```{r}
df_raw <- read.csv("kc_house_data.csv")
str(df_raw)
summary(df_raw$price)
```

# Data Cleaning and Preprocessing
We convert appropriate variables to factors, create a log-transformed response, and remove identifier-type fields that are not useful as predictors.
```{r}
df <- df_raw %>%
  mutate(
    zipcode = factor(zipcode),
    waterfront = factor(waterfront),
    view = factor(view),
    condition = factor(condition),
    grade = factor(grade),
    log_price = log(price)
  )

# Drop identifier / raw date fields if present
drop_cols <- intersect(names(df), c("id", "date"))
df_model <- df %>% select(-all_of(drop_cols))

glimpse(df_model)
```

# Train/Test Split
We use an 80/20 train/test split. All models are trained on the same training set and evaluated on the same test set for a fair comparison.
```{r}
set.seed(5703)
idx <- createDataPartition(df_model$log_price, p = 0.80, list = FALSE)

train_df <- df_model[idx, , drop = FALSE]
test_df  <- df_model[-idx, , drop = FALSE]

cat("Training rows:", nrow(train_df), "\n")
cat("Test rows:", nrow(test_df), "\n")

# Encode zipcode using training data only
zip_medians <- train_df %>%
  group_by(zipcode) %>%
  summarize(zipcode_log_median = median(log_price), .groups = "drop")

train_df <- train_df %>%
  left_join(zip_medians, by = "zipcode")

test_df <- test_df %>%
  left_join(zip_medians, by = "zipcode")

overall_train_log_median <- median(train_df$log_price)

test_df <- test_df %>%
  mutate(
    zipcode_log_median = if_else(
      is.na(zipcode_log_median),
      overall_train_log_median,
      zipcode_log_median
    )
  )

# Remove original zipcode after encoding
train_df <- train_df %>% select(-zipcode)
test_df  <- test_df %>% select(-zipcode)

glimpse(train_df)
glimpse(test_df)
```

# Model Matrices for Bagging and Random Forest
For randomForest(), we create predictor matrices and response vectors. We exclude both price and log_price from predictors, and use log_price as the response.
```{r}
x_train <- train_df %>% select(-price, -log_price)
x_test  <- test_df %>% select(-price, -log_price)

y_train <- train_df$log_price
y_test  <- test_df$log_price

p <- ncol(x_train)

cat("Number of predictors used:", p, "\n")
```

# Baseline Model: Single Regression Tree
A single regression tree is included only as a benchmark. It is not one of the required ensemble methods, but it helps show whether the ensembles improve predictive performance.
```{r}
tree_model <- rpart(
  log_price ~ . - price,
  data = train_df,
  method = "anova",
  control = rpart.control(cp = 0.0001)
)

printcp(tree_model)
plotcp(tree_model)

best_cp <- tree_model$cptable[
  which.min(tree_model$cptable[, "xerror"]),
  "CP"
]

cat("Best cp selected from cross-validation table:", best_cp, "\n")

pruned_tree <- prune(tree_model, cp = best_cp)

rpart.plot(pruned_tree, main = "Pruned Regression Tree")

pred_log_tree <- predict(pruned_tree, newdata = test_df)

tree_test_metrics <- print_price_metrics(
  "Decision Tree Test Metrics:",
  pred_log_tree,
  test_df$price
)
```

# Bagging
Bagging fits many trees using bootstrap samples and averages the predictions. For bagging, the number of variables tried at each split is set equal to the total number of predictors, so mtry = p.
```{r}
set.seed(5703)

bagging_model <- randomForest(
  x = x_train,
  y = y_train,
  mtry = p,
  ntree = 300,
  importance = TRUE
)

bagging_model
varImpPlot(bagging_model, main = "Bagging Variable Importance")

pred_log_bag <- predict(bagging_model, newdata = x_test)

bagging_test_metrics <- print_price_metrics(
  "Bagging Test Metrics:",
  pred_log_bag,
  test_df$price
)
```

# Random Forest
Random forest improves on bagging by considering a random subset of predictors at each split. For regression, a common default is mtry = floor(sqrt(p)) or sometimes p/3. Here we tune a small grid and choose the best OOB performance.
```{r}
set.seed(5703)

rf_grid <- c(
  max(2, floor(sqrt(p))),
  max(2, floor(p / 3)),
  max(2, floor(p / 2))
)

rf_grid <- unique(rf_grid)

rf_tuning_results <- map_dfr(rf_grid, function(m) {
  fit <- randomForest(
    x = x_train,
    y = y_train,
    mtry = m,
    ntree = 300,
    importance = TRUE
  )
  
  tibble(
    mtry = m,
    OOB_MSE = tail(fit$mse, 1)
  )
})

rf_tuning_results %>%
  arrange(OOB_MSE) %>%
  kable(caption = "Random Forest Tuning Results")

best_mtry <- rf_tuning_results %>%
  arrange(OOB_MSE) %>%
  slice(1) %>%
  pull(mtry)

cat("Best mtry for random forest:", best_mtry, "\n")

set.seed(5703)
rf_model <- randomForest(
  x = x_train,
  y = y_train,
  mtry = best_mtry,
  ntree = 300,
  importance = TRUE
)

rf_model
varImpPlot(rf_model, main = "Random Forest Variable Importance")

pred_log_rf <- predict(rf_model, newdata = x_test)

rf_test_metrics <- print_price_metrics(
  "Random Forest Test Metrics:",
  pred_log_rf,
  test_df$price
)
```

# Boosting
Boosting builds trees sequentially, with each new tree focusing on errors made by the previous ones. We use gbm() with Gaussian loss because this is a regression problem.
```{r}
set.seed(5703)

boost_model <- gbm(
  log_price ~ . - price,
  data = train_df,
  distribution = "gaussian",
  n.trees = 3000,
  interaction.depth = 3,
  shrinkage = 0.01,
  n.minobsinnode = 10,
  cv.folds = 5,
  verbose = FALSE
)

best_iter <- gbm.perf(boost_model, method = "cv", plot.it = TRUE)
cat("Best number of trees from CV:", best_iter, "\n")

pred_log_boost <- predict(
  boost_model,
  newdata = test_df,
  n.trees = best_iter
)

boost_test_metrics <- print_price_metrics(
  "Boosting Test Metrics:",
  pred_log_boost,
  test_df$price
)
```

# Model Comparison
This table compares the benchmark tree and the three required ensemble methods using the same held-out test set.
```{r}
comparison_table <- bind_rows(
  tibble(
    Method = "Decision Tree",
    Tuning = paste0("cp = ", round(best_cp, 5)),
    RMSE = tree_test_metrics$RMSE,
    MAE = tree_test_metrics$MAE,
    MSE = tree_test_metrics$MSE,
    R2 = tree_test_metrics$R2
  ),
  tibble(
    Method = "Bagging",
    Tuning = paste0("ntree = 300, mtry = ", p),
    RMSE = bagging_test_metrics$RMSE,
    MAE = bagging_test_metrics$MAE,
    MSE = bagging_test_metrics$MSE,
    R2 = bagging_test_metrics$R2
  ),
  tibble(
    Method = "Random Forest",
    Tuning = paste0("ntree = 300, mtry = ", best_mtry),
    RMSE = rf_test_metrics$RMSE,
    MAE = rf_test_metrics$MAE,
    MSE = rf_test_metrics$MSE,
    R2 = rf_test_metrics$R2
  ),
  tibble(
    Method = "Boosting",
    Tuning = paste0("n.trees = ", best_iter, ", depth = 3, shrinkage = 0.01"),
    RMSE = boost_test_metrics$RMSE,
    MAE = boost_test_metrics$MAE,
    MSE = boost_test_metrics$MSE,
    R2 = boost_test_metrics$R2
  )
) %>%
  arrange(RMSE)

kable(
  comparison_table,
  digits = 4,
  caption = "Comparison of Decision Tree, Bagging, Random Forest, and Boosting"
)
```

# Variable Importance
These plots help interpret which predictors matter most for the ensemble tree methods.
```{r}
importance_rf <- importance(rf_model)
importance_bag <- importance(bagging_model)

rf_importance_df <- data.frame(
  Variable = rownames(importance_rf),
  Importance = importance_rf[, 1]
) %>%
  arrange(desc(Importance)) %>%
  slice_head(n = 15)

bag_importance_df <- data.frame(
  Variable = rownames(importance_bag),
  Importance = importance_bag[, 1]
) %>%
  arrange(desc(Importance)) %>%
  slice_head(n = 15)

kable(rf_importance_df, digits = 4, caption = "Top 15 Random Forest Importance Values")
kable(bag_importance_df, digits = 4, caption = "Top 15 Bagging Importance Values")
```

# Interpretation of Results
Based on the comparison table, the best model should be identified primarily by the lowest RMSE and MAE, along with the highest R^2 on the test set.
In general:
The decision tree serves as a simple benchmark and is usually less accurate than the ensemble methods.
Bagging often improves over a single tree by reducing variance.
Random forest often improves further by decorrelating trees through random predictor selection.
Boosting can perform very well because it builds trees sequentially to correct earlier errors.
The final recommendation should be based on the model that performs best on the held-out test set while remaining interpretable and appropriately tuned.

# Final Conclusion
This project applied the three required ensemble tree methods from Chapter 8 to a house price prediction problem:
Bagging
Random Forest
Boosting
All models were trained on the same training set and evaluated on the same test set using regression metrics:
RMSE
MAE
MSE
R^2
This provides a fair comparison of predictive performance. The final selected model should be the one with the strongest test-set performance, supported by a clear interpretation of why its results make sense for this dataset.