---
title: "Final Project Step 3: Ensemble Tree Methods for House Price Prediction"
author: "Group 6: Cristhian Cruz, Julian McKinley, Luisa Ribeiro, Matheus Lucena"
date: "2026-04-13"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  results = "hold"
)
```

# Problem Statement
The goal of this project is to predict house sale price using housing characteristics from the King County housing dataset. This is a regression problem, since the response variable is continuous.
This problem matters because accurate house price prediction is useful in real estate valuation, pricing strategy, and market analysis. In this final step of the project, we focus on ensemble tree methods from Chapter 8 
and compare:

Bagging
Random Forest
Boosting

A single regression tree is also included as a benchmark, but the main emphasis is on the three required ensemble methods.

# Packages

```{r}
suppressPackageStartupMessages({
  library(tidyverse)
  library(caret)
  library(rpart)
  library(rpart.plot)
  library(gbm)
  library(knitr)
  library(ranger)
  library(parallel)
  library(pdp)
})

N_CORES <- max(1, detectCores() - 1)   # leave 1 core free
```

```{r}
N_CORES
```

# Helper Functions
```{r}
set.seed(5703)

price_metrics <- function(pred_log, actual_price) {
  pred_price <- exp(as.numeric(pred_log))
  mse <- mean((actual_price - pred_price)^2)
  rmse <- sqrt(mse)
  mae <- mean(abs(actual_price - pred_price))
  r2 <- 1 - sum((actual_price - pred_price)^2) /
    sum((actual_price - mean(actual_price))^2)
  
  tibble(
    RMSE = rmse,
    MAE = mae,
    MSE = mse,
    R2 = r2
  )
}

print_price_metrics <- function(model_name, pred_log, actual_price) {
  out <- price_metrics(pred_log, actual_price)
  cat(model_name, "\n")
  cat("RMSE:", round(out$RMSE, 3), "\n")
  cat("MAE :", round(out$MAE, 3), "\n")
  cat("MSE :", round(out$MSE, 3), "\n")
  cat("R^2 :", round(out$R2, 6), "\n\n")
  invisible(out)
}

extract_oob_mse <- function(rf_fit) {
  rf_fit$prediction.error
}

extract_oob_rsq <- function(rf_fit) {
  rf_fit$r.squared
}
```

# Data Description and Loading
The dataset used is kc_house_data.csv. The response variable is price. Because house prices are strongly right-skewed, we model log(price) and then transform predictions back to the original price scale for evaluation.
```{r}
df_raw <- read.csv("kc_house_data.csv")
str(df_raw)
summary(df_raw$price)
```

# Data Cleaning and Preprocessing
We convert appropriate variables to factors, create a log-transformed response, and remove identifier-type fields that are not useful as predictors.
```{r}
df <- df_raw %>%
  mutate(
    # Convert to proper datetime
    date = as.POSIXct(date, format = "%Y%m%dT%H%M%S"),
    
    # Extract useful features
    sale_year  = as.numeric(format(date, "%Y")),
    sale_month = as.numeric(format(date, "%m")),
    sale_day   = as.numeric(format(date, "%d")),
    sale_dow   = as.numeric(format(date, "%u")),  # day of week (1–7)
    
    # Optional: continuous time feature
    sale_time_numeric = as.numeric(date),
    
    # Existing transformations
    zipcode = factor(zipcode),
    waterfront = factor(waterfront),
    view = factor(view),
    condition = factor(condition),
    grade = factor(grade),
    log_price = log(price)
  )

# Drop identifier / raw date fields if present
drop_cols <- intersect(names(df), c("id", "date"))
df_model <- df %>% select(-all_of(drop_cols))

glimpse(df_model)
```

# Train/Test Split
We use an 80/20 train/test split. All models are trained on the same training set and evaluated on the same test set for a fair comparison.
```{r}
set.seed(5703)
idx <- createDataPartition(df_model$log_price, p = 0.80, list = FALSE)

train_df <- df_model[idx, , drop = FALSE]
test_df  <- df_model[-idx, , drop = FALSE]

cat("Training rows:", nrow(train_df), "\n")
cat("Test rows:", nrow(test_df), "\n")

# Encode zipcode using training data only
zip_medians <- train_df %>%
  group_by(zipcode) %>%
  summarize(zipcode_log_median = median(log_price), .groups = "drop")

train_df <- train_df %>%
  left_join(zip_medians, by = "zipcode")

test_df <- test_df %>%
  left_join(zip_medians, by = "zipcode")

overall_train_log_median <- median(train_df$log_price)

test_df <- test_df %>%
  mutate(
    zipcode_log_median = if_else(
      is.na(zipcode_log_median),
      overall_train_log_median,
      zipcode_log_median
    )
  )

# Remove original zipcode after encoding
train_df <- train_df %>% select(-zipcode)
test_df  <- test_df %>% select(-zipcode)

glimpse(train_df)
glimpse(test_df)
```

# Model Matrices for Bagging and Random Forest
For ranger(), we create predictor matrices and response vectors. We exclude both price and log_price from predictors, and use log_price as the response.
```{r}
x_train <- train_df %>% select(-price, -log_price)
x_test  <- test_df %>% select(-price, -log_price)

y_train <- train_df$log_price
y_test  <- test_df$log_price

p <- ncol(x_train)

cat("Number of predictors used:", p, "\n")
```

# Baseline Model: Single Regression Tree
A single regression tree is included only as a benchmark. It is not one of the required ensemble methods, but it helps show whether the ensembles improve predictive performance.
```{r}
tree_model <- rpart(
  log_price ~ . - price,
  data = train_df,
  method = "anova",
  control = rpart.control(cp = 0.0001)
)

printcp(tree_model)
plotcp(tree_model)

tree_cp_df <- as.data.frame(tree_model$cptable)

ggplot(tree_cp_df, aes(x = nsplit, y = xerror)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Decision Tree Cross-Validation Error vs Tree Size",
    x = "Number of Splits",
    y = "Cross-Validated Relative Error"
  ) +
  theme_minimal()

best_cp <- tree_model$cptable[
  which.min(tree_model$cptable[, "xerror"]),
  "CP"
]

cat("Best cp selected from cross-validation table:", best_cp, "\n")

pruned_tree <- prune(tree_model, cp = best_cp)

rpart.plot(pruned_tree, main = "Pruned Regression Tree")

pred_log_tree <- predict(pruned_tree, newdata = test_df)

tree_test_metrics <- print_price_metrics(
  "Decision Tree Test Metrics:",
  pred_log_tree,
  test_df$price
)
```
```{r}
smaller_tree <- prune(pruned_tree, cp = 0.01)
rpart.plot(smaller_tree)
```


# Bagging
Bagging fits many trees using bootstrap samples and averages the predictions. For bagging, the number of variables tried at each split is set equal to the total number of predictors, so mtry = p.
```{r}
set.seed(5703)

bag_ntree_grid <- c(300, 500, 1000)

bagging_tuning_results <- map_dfr(bag_ntree_grid, function(nt) {
  fit <- ranger(
    x = x_train,
    y = y_train,
    mtry = p,
    num.trees = nt,
    importance = "impurity"
  )
  
  tibble(
    ntree = nt,
    mtry = p,
    OOB_MSE = extract_oob_mse(fit),
    OOB_R2 = extract_oob_rsq(fit)
  )
})

ggplot(bagging_tuning_results, aes(x = ntree, y = OOB_MSE)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Bagging OOB Error vs Number of Trees",
    x = "Number of Trees",
    y = "OOB MSE"
  ) +
  theme_minimal()

kable(
  bagging_tuning_results,
  digits = 5,
  caption = "Bagging tuning results based on out-of-bag error"
)

best_bag_ntree <- bagging_tuning_results %>%
  arrange(OOB_MSE) %>%
  slice(1) %>%
  pull(ntree)

cat("Best ntree for bagging:", best_bag_ntree, "\n")

bagging_model <- ranger(
  x = x_train,
  y = y_train,
  mtry = p,
  num.trees = best_bag_ntree,
  importance = "impurity"
)

bagging_model
bag_importance_values <- sort(importance(bagging_model), decreasing = TRUE)
barplot(
  head(bag_importance_values, 15),
  las = 2,
  cex.names = 0.7,
  main = "Top 15 Bagging Variable Importance",
  ylab = "Importance"
)

pred_log_bag <- predict(bagging_model, data = x_test)$predictions

bagging_test_metrics <- print_price_metrics(
  "Bagging Test Metrics:",
  pred_log_bag,
  test_df$price
)
```

# Random Forest
Random forest improves on bagging by considering a random subset of predictors at each split. Instead of a very small tuning grid, we tune both mtry and ntree using out-of-bag performance and then fit the final model with the best settings.
```{r}
set.seed(5703)

rf_mtry_grid <- unique(c(
  2,
  max(2, floor(sqrt(p))),
  max(2, floor(p / 4)),
  max(2, floor(p / 3)),
  max(2, floor(p / 2)),
  p
))

rf_ntree_grid <- c(300, 500, 1000)

rf_grid <- expand.grid(
  mtry = rf_mtry_grid,
  ntree = rf_ntree_grid
) %>%
  as_tibble() %>%
  arrange(mtry, ntree)

rf_tuning_results <- pmap_dfr(rf_grid, function(mtry, ntree) {
  fit <- ranger(
    x = x_train,
    y = y_train,
    mtry = mtry,
    num.trees = ntree,
    importance = "impurity"
  )
  
  tibble(
    mtry = mtry,
    ntree = ntree,
    OOB_MSE = extract_oob_mse(fit),
    OOB_R2 = extract_oob_rsq(fit)
  )
})

rf_tuning_results %>%
  arrange(OOB_MSE) %>%
  kable(digits = 5, caption = "Random Forest Tuning Results")

best_rf_mtry_plot <- rf_tuning_results %>%
  arrange(OOB_MSE) %>%
  slice(1) %>%
  pull(mtry)

rf_ntree_plot_df <- rf_tuning_results %>%
  filter(mtry == best_rf_mtry_plot)

ggplot(rf_ntree_plot_df, aes(x = ntree, y = OOB_MSE)) +
  geom_line() +
  geom_point() +
  labs(
    title = paste("Random Forest OOB Error vs Number of Trees (mtry =", best_rf_mtry_plot, ")"),
    x = "Number of Trees",
    y = "OOB MSE"
  ) +
  theme_minimal()

best_rf_params <- rf_tuning_results %>%
  arrange(OOB_MSE) %>%
  slice(1)

best_mtry <- best_rf_params$mtry
best_rf_ntree <- best_rf_params$ntree

cat("Best mtry for random forest:", best_mtry, "\n")
cat("Best ntree for random forest:", best_rf_ntree, "\n")

set.seed(5703)
rf_model <- ranger(
  x = x_train,
  y = y_train,
  mtry = best_mtry,
  num.trees = best_rf_ntree,
  importance = "impurity"
)

rf_model
rf_importance_values <- sort(importance(rf_model), decreasing = TRUE)
barplot(
  head(rf_importance_values, 15),
  las = 2,
  cex.names = 0.7,
  main = "Top 15 Random Forest Variable Importance",
  ylab = "Importance"
)

pred_log_rf <- predict(rf_model, data = x_test)$predictions

rf_test_metrics <- print_price_metrics(
  "Random Forest Test Metrics:",
  pred_log_rf,
  test_df$price
)
```

# Boosting
Boosting builds trees sequentially, with each new tree focusing on errors made by the previous ones. Here we tune interaction depth, shrinkage, and minimum node size, then choose the best number of trees by cross-validation for each combination.
```{r}
set.seed(5703)

boost_grid <- expand.grid(
  interaction.depth = c(1, 3, 5),
  shrinkage = c(0.005, 0.01),
  n.minobsinnode = c(5, 10),
  KEEP.OUT.ATTRS = FALSE
) %>%
  as_tibble()

boost_tuning_results <- pmap_dfr(boost_grid, function(interaction.depth, shrinkage, n.minobsinnode) {
  fit <- gbm(
    log_price ~ . - price,
    data = train_df,
    distribution = "gaussian",
    n.trees = 4000,
    interaction.depth = interaction.depth,
    shrinkage = shrinkage,
    n.minobsinnode = n.minobsinnode,
    cv.folds = 5,
    n.cores = N_CORES,
    bag.fraction = 0.5,
    verbose = FALSE
  )
  
  best_iter <- gbm.perf(fit, method = "cv", plot.it = FALSE)
  best_cv_error <- fit$cv.error[best_iter]
  
  tibble(
    interaction.depth = interaction.depth,
    shrinkage = shrinkage,
    n.minobsinnode = n.minobsinnode,
    best_iter = best_iter,
    CV_Error = best_cv_error
  )
})

boost_tuning_results %>%
  arrange(CV_Error) %>%
  kable(digits = 5, caption = "Boosting tuning results based on 5-fold CV")

best_boost_params <- boost_tuning_results %>%
  arrange(CV_Error) %>%
  slice(1)

cat("Best boosting depth:", best_boost_params$interaction.depth, "\n")
cat("Best boosting shrinkage:", best_boost_params$shrinkage, "\n")
cat("Best boosting n.minobsinnode:", best_boost_params$n.minobsinnode, "\n")
cat("Best boosting number of trees:", best_boost_params$best_iter, "\n")

boost_model <- gbm(
  log_price ~ . - price,
  data = train_df,
  distribution = "gaussian",
  n.trees = best_boost_params$best_iter,
  interaction.depth = best_boost_params$interaction.depth,
  shrinkage = best_boost_params$shrinkage,
  n.minobsinnode = best_boost_params$n.minobsinnode,
  cv.folds = 5,
  n.cores = N_CORES,
  bag.fraction = 0.5,
  verbose = FALSE
)

boost_cv_df <- data.frame(
  Trees = seq_along(boost_model$cv.error),
  CV_Error = boost_model$cv.error
)

best_iter_final <- gbm.perf(boost_model, method = "cv", plot.it = FALSE)

ggplot(boost_cv_df, aes(x = Trees, y = CV_Error)) +
  geom_line() +
  geom_vline(xintercept = best_iter_final, linetype = "dashed") +
  labs(
    title = "Boosting Cross-Validation Error vs Number of Trees",
    x = "Number of Trees",
    y = "CV Error"
  ) +
  theme_minimal()

boost_importance_values <- summary(boost_model, plotit = FALSE)
gbm_importance_values <- boost_importance_values$rel.inf
names(gbm_importance_values) <- boost_importance_values$var
barplot(
  head(gbm_importance_values, 15),
  las = 2,
  cex.names = 0.7,
  main = "Top 15 boosting Variable Importance",
  ylab = "Relative Influence"
)

pred_log_boost <- predict(
  boost_model,
  newdata = test_df,
  n.trees = best_boost_params$best_iter
)

boost_test_metrics <- print_price_metrics(
  "Boosting Test Metrics:",
  pred_log_boost,
  test_df$price
)

# Residual Analysis for Boosting 
actual_price <- test_df$price
pred_price <- exp(pred_log_boost)
residuals <- actual_price - pred_price

# Residual Plot
plot(pred_price, residuals,
     xlab = "Predicted Price",
     ylab = "Residuals",
     main = "Residual Plot (Boosting Model)")
abline(h = 0, lwd = 2)

# Actual vs Predicted
plot(actual_price, pred_price,
     xlab = "Actual Price",
     ylab = "Predicted Price",
     main = "Actual vs Predicted (Boosting Model)")
abline(0, 1, lwd = 2)

# Residual Distribution
hist(residuals,
     breaks = 30,
     main = "Residual Distribution (Boosting Model)",
     xlab = "Residuals")
```

# Partial Dependence Plot
```{r}
boost_zip_pdp <- partial(
  object = boost_model,
  pred.var = "zipcode_log_median",
  train = x_train,
  grid.resolution = 20,
  n.trees = 4000
)

autoplot(boost_zip_pdp) +
  labs(
    title = "Partial Dependence Plot for Zipcode Encoding",
    x = "Zipcode Log Median",
    y = "Predicted log(Price)"
  ) +
  theme_minimal()
```

# Model Comparison
This table compares the benchmark tree and the three required ensemble methods using the same held-out test set.
```{r}
comparison_table <- bind_rows(
  tibble(
    Method = "Decision Tree",
    Tuning = paste0("cp = ", round(best_cp, 5)),
    RMSE = tree_test_metrics$RMSE,
    MAE = tree_test_metrics$MAE,
    MSE = tree_test_metrics$MSE,
    R2 = tree_test_metrics$R2
  ),
  tibble(
    Method = "Bagging",
    Tuning = paste0("ntree = ", best_bag_ntree, ", mtry = ", p),
    RMSE = bagging_test_metrics$RMSE,
    MAE = bagging_test_metrics$MAE,
    MSE = bagging_test_metrics$MSE,
    R2 = bagging_test_metrics$R2
  ),
  tibble(
    Method = "Random Forest",
    Tuning = paste0("ntree = ", best_rf_ntree, ", mtry = ", best_mtry),
    RMSE = rf_test_metrics$RMSE,
    MAE = rf_test_metrics$MAE,
    MSE = rf_test_metrics$MSE,
    R2 = rf_test_metrics$R2
  ),
  tibble(
    Method = "Boosting",
    Tuning = paste0(
      "n.trees = ", best_boost_params$best_iter,
      ", depth = ", best_boost_params$interaction.depth,
      ", shrinkage = ", best_boost_params$shrinkage,
      ", n.minobsinnode = ", best_boost_params$n.minobsinnode
    ),
    RMSE = boost_test_metrics$RMSE,
    MAE = boost_test_metrics$MAE,
    MSE = boost_test_metrics$MSE,
    R2 = boost_test_metrics$R2
  )
) %>%
  arrange(RMSE)

kable(
  comparison_table,
  digits = 4,
  caption = "Comparison of Decision Tree, Bagging, Random Forest, and Boosting"
)
```

# Variable Importance
These plots help interpret which predictors matter most for the ensemble tree methods.
```{r}

boost_importance <- summary(boost_model, plotit = FALSE)
importance_rf <- importance(rf_model)
importance_bag <- importance(bagging_model)
importance_tree <- pruned_tree$variable.importance

boost_importance_df <- boost_importance %>%
  select(Variable = var, Importance = rel.inf) %>%
  arrange(desc(Importance)) %>%
  slice_head(n = 15)

rf_importance_df <- data.frame(
  Variable = names(importance_rf),
  Importance = as.numeric(importance_rf)
) %>%
  arrange(desc(Importance)) %>%
  slice_head(n = 15)

bag_importance_df <- data.frame(
  Variable = names(importance_bag),
  Importance = as.numeric(importance_bag)
) %>%
  arrange(desc(Importance)) %>%
  slice_head(n = 15)

tree_importance_df <- data.frame(
  Variable = names(importance_tree),
  Importance = as.numeric(importance_tree)
) %>%
  arrange(desc(Importance)) %>%
  slice_head(n = 15)

kable(boost_importance_df, digits = 4, caption = "Top 15 Boosting Importance Values")
kable(rf_importance_df, digits = 4, caption = "Top 15 Random Forest Importance Values")
kable(bag_importance_df, digits = 4, caption = "Top 15 Bagging Importance Values")
kable(tree_importance_df, digits = 4, caption = "Top 15 Decision Tree Importance Values")
```
