---
title: "Project2 Data Mining"
author: "Group 6: Cristhian Cruz, Julian McKinley, Luisa Ribeiro, Matheus Lucena"
date: "2026-03-03"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      results = "hold",
                      message = FALSE,
                      warning = FALSE)
```

# Project part 2

## Setup and data cleaning
```{r}
suppressPackageStartupMessages({
  library(tidyverse)
  library(caret)
  library(MASS)
  library(leaps)
  library(glmnet)
  library(dplyr)
})

set.seed(5703)
df_raw <- read.csv("kc_house_data.csv")
```

```{r}
# Convert likely categorical predictors
df <- df_raw %>%
  mutate(
    zipcode = if ("zipcode" %in% names(.)) as.factor(zipcode) else NULL,
    waterfront = if ("waterfront" %in% names(.)) as.factor(waterfront) else NULL,
    view = if ("view" %in% names(.)) as.factor(view) else NULL,
    condition = if ("condition" %in% names(.)) as.factor(condition) else NULL,
    grade = if ("grade" %in% names(.)) as.factor(grade) else NULL
  )

# We will be using log(price) for final model
df <- df %>%
  mutate(log_price = log(price)) %>%
  dplyr::select(-price)
```

```{r}
# Check for variables with perfect multicollinearity 
alias_obj <- alias(lm(log_price ~ ., data = df))
redundant_cols <- rownames(alias_obj$Complete)
redundant_cols
```

```{r}
# Drop ID, Date + redundant variables from modeling
drop_cols <- intersect(names(df), c("id", "date", redundant_cols))
df_model <- dplyr::select(df, -dplyr::all_of(drop_cols))
```

```{r}
# Train/Test Split
set.seed(5703)
idx <- createDataPartition(df_model$log_price, p = 0.80, list = FALSE)
train_df <- df_model[idx, , drop = FALSE]
test_df <- df_model[-idx, , drop = FALSE]

cat("Train rows:", nrow(train_df), "\nTest rows:", nrow(test_df), "\n")
```

## Part 1: Model selection methods

### A. Best Subset Selection

`zipcode` is not included in best subset selection because it has 70 levels once dummy encoded. This results in too many predictors for `regsubsets` to handle.

```{r}
# get the number of predictors after dummy encoding categorical predictors
p <- ncol(model.matrix(~ . - log_price -zipcode, data=train_df))

best_fit <- regsubsets(log_price ~ . - zipcode, data = train_df, nvmax = p)

best_summary <-summary(best_fit)

# View metrics
cat("Regsubsets Metrics\nAdjusted R^2:\n")
best_summary$adjr2
cat("\nCP:\n")
best_summary$cp
cat("\nBIC:\n")
best_summary$bic

# Identify optimal model size for each criterion
best_adjr2 <- which.max(best_summary$adjr2)
best_cp    <- which.min(best_summary$cp)
best_bic   <- which.min(best_summary$bic)

cat("\nIndexes of best subset for each metric:\nR^2: ", best_adjr2, "\n")
cat("Cp: ", best_cp, "\n")
cat("BIC: ", best_bic, "\n")
```

```{r}
# Plots
par(mfrow = c(1,3))

plot(best_summary$adjr2, xlab = "Number of predictors", ylab = "Adjusted R2", type = "b")
points(best_adjr2, best_summary$adjr2[best_adjr2], col = "red", pch = 19)

plot(best_summary$cp, xlab = "Number of predictors", ylab = "Cp", type = "b")
points(best_cp, best_summary$cp[best_cp], col = "red", pch = 19)

plot(best_summary$bic, xlab = "Number of predictors", ylab = "BIC", type = "b")
points(best_bic, best_summary$bic[best_bic], col = "red", pch = 19)

par(mfrow = c(1,1))
```

Refitting the model using the best BIC subset. We're using the all-or-nothing rule for categorical values. If one of the categorical levels is selected, the whole categorical variable is selected in the final model.

```{r}
best_bic_coefs <- coef(best_fit, best_bic)
# Extract predictor names (remove intercept)
selected_vars <- names(best_bic_coefs)[-1]
all_original_vars <- colnames(train_df)
# Selected predictors might be a dummy-coded variable, so we need to map it back to its original variable name.
final_selection <- all_original_vars[sapply(all_original_vars, function(v) any(grepl(v, selected_vars)))]
cat("Best variable subset selected by regsubsets:\n")
selected_vars
cat("\n\nVariables selected for final model:\n")
final_selection
```

```{r}
# Build formula dynamically
formula_bss <- as.formula(
  paste("log_price ~", paste(final_selection, collapse = " + "))
)

# Refit model using training data
bss_model <- lm(formula_bss, data = train_df)
summary(bss_model)

# Predictions on test set
pred_test <- predict(bss_model, newdata = test_df)

# Test RMSE
rmse_bss <- sqrt(mean((test_df$log_price - pred_test)^2))

# Test R-squared
r2_bss <- 1 - sum((test_df$log_price - pred_test)^2) /
  sum((test_df$log_price - mean(test_df$log_price))^2)

cat("Test set metric of model with best subset selection\n")
cat("RMSE: ", rmse_bss, "\nR^2: ", r2_bss)
```

### B. Stepwise Selection(Forward, Backward or Both)
```{r}
full_model <- lm(log_price ~ ., data = train_df)

step_model <- step(full_model, direction = "both", trace = FALSE)
summary(step_model)

#top 10 significant predictors
top_vars <- broom::tidy(step_model) %>%
  filter(p.value < 0.01) %>%
  arrange(desc(abs(statistic))) %>%
  head(10)

top_vars

# Predict on test set
pred_step <- predict(step_model, newdata = test_df)

# Test RMSE
rmse_step <- sqrt(mean((test_df$log_price - pred_step)^2))

# Test R^2
r2_step <- 1 - sum((test_df$log_price - pred_step)^2) /
           sum((test_df$log_price - mean(test_df$log_price))^2)

cat("Stepwise Test RMSE:", rmse_step, "\n")
cat("Stepwise Test R^2:", r2_step, "\n")
```

## Part 2: Regularization and 5-fold cross validation

our outcome is continuous so we will be using ridge and lasso
```{r}
# exclude price and log_price from predictors
x_train <- model.matrix(log_price ~ . - log_price, data = train_df)[, -1]
y_train <- train_df$log_price

x_test <- model.matrix(log_price ~ . - log_price, data = test_df)[, -1]
y_test <- test_df$log_price

# 5-fold CV for Ridge Regression
set.seed(5703)
cv_ridge <- cv.glmnet(x_train, y_train, alpha = 0, nfolds = 5)
plot(cv_ridge)
title("Ridge Regression CV Error Curve")

# Optimal lambdas
cv_ridge$lambda.min
cv_ridge$lambda.1se

# Ridge coefficients at lambda.min
coef(cv_ridge, s = "lambda.min")

ridge_pred <- predict(cv_ridge, s = "lambda.min", newx = x_test)
rmse_ridge <- sqrt(mean((y_test - ridge_pred)^2))
cat("Ridge Test RMSE:", rmse_ridge)

# 5-fold CV for LASSO Regression
set.seed(5703)
cv_lasso <- cv.glmnet(x_train, y_train, alpha = 1, nfolds = 5)
plot(cv_lasso)
title("LASSO Regression CV Error Curve")

# Optimal lambdas
cv_lasso$lambda.min
cv_lasso$lambda.1se

# LASSO coefficients at lambda.min
lasso_coef <- coef(cv_lasso, s = "lambda.min")
lasso_coef

lasso_pred <- predict(cv_lasso, s = "lambda.min", newx = x_test)
rmse_lasso <- sqrt(mean((y_test - lasso_pred)^2))
cat("LASSO Test RMSE:", rmse_lasso)

# Which predictors are zero
zero_predictors <- rownames(lasso_coef)[lasso_coef[, 1] == 0]
zero_predictors

model_results <- data.frame(
  Model = c("Best Subset (BIC)", "Stepwise", "Ridge", "LASSO"),
  RMSE = c(rmse_bss, rmse_step, rmse_ridge, rmse_lasso)
)

model_results
```

## Part 3: Model performance evaluation and final choice 
In the written report.
