---
title: "STA 5703 Project 2 - Submission-Ready R Markdown"
author: "Group 6: Cristhian Cruz, Julian McKinley, Luisa Ribeiro, Matheus Lucena"
date: "2026-03-11"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  results = "hold"
)
```

# Project goal

This file is organized to match the Project Presentation #2 rubric. The response variable is **house sale price**, and because price is strongly right-skewed, the regression models are fit on **log(price)** and then evaluated back on the original **price** scale using RMSE and test-set $R^2$.

The main goals are:

1.  compare **best subset selection** and **stepwise selection**;
2.  fit **ridge** and **lasso** using **5-fold cross-validation**;
3.  compare model performance on a held-out test set;
4.  justify a final model choice while discussing complexity and overfitting.

# 1. Setup and data preparation

```{r packages}
suppressPackageStartupMessages({
  library(tidyverse)
  library(caret)
  library(MASS)
  library(leaps)
  library(glmnet)
  library(broom)
  library(knitr)
})

set.seed(5703)
```

```{r load-data}
df_raw <- read.csv("kc_house_data.csv")
str(df_raw)
```

```{r preprocess-data}
# Convert categorical predictors to factors
# and create the log-transformed outcome used in modeling.
df <- df_raw %>%
  mutate(
    zipcode = factor(zipcode),
    waterfront = factor(waterfront),
    view = factor(view),
    condition = factor(condition),
    grade = factor(grade),
    log_price = log(price)
  )
```

```{r alias-check}
# Check for perfect multicollinearity before modeling.
alias_obj <- alias(lm(log_price ~ ., data = df))
redundant_cols <- rownames(alias_obj$Complete)
redundant_cols
```

```{r drop-columns}
# Drop identifiers and any aliased columns.
drop_cols <- intersect(names(df), c("id", "date", redundant_cols))
df_model <- df %>% dplyr::select(-dplyr::all_of(drop_cols))

names(df_model)
```

```{r split-data}
# 80/20 train-test split
idx <- createDataPartition(df_model$log_price, p = 0.80, list = FALSE)
train_df <- df_model[idx, , drop = FALSE]
test_df  <- df_model[-idx, , drop = FALSE]

cat("Train rows:", nrow(train_df), "\n")
cat("Test rows:", nrow(test_df), "\n")
```

## 1.1 Target encoding for zipcode

`zipcode` has many levels, so for subset-selection methods we replace it with a **training-only target-encoded feature** named `zipcode_log_median`. This preserves location information while avoiding a very large set of dummy variables. The encoding is learned from the training set only to avoid leakage.

```{r zipcode-encoding}
zip_medians <- train_df %>%
  group_by(zipcode) %>%
  summarize(zipcode_log_median = median(log_price), .groups = "drop")

train_df <- train_df %>% left_join(zip_medians, by = "zipcode")
test_df  <- test_df  %>% left_join(zip_medians, by = "zipcode")

overall_train_log_median <- median(train_df$log_price)

test_df <- test_df %>%
  mutate(
    zipcode_log_median = if_else(
      is.na(zipcode_log_median),
      overall_train_log_median,
      zipcode_log_median
    )
  )

summary(train_df$zipcode_log_median)
summary(test_df$zipcode_log_median)
```

```{r remove-original-zipcode}
# Remove the original zipcode after encoding.
df_model <- df %>% dplyr::select(-dplyr::all_of(drop_cols))
train_df <- train_df %>% dplyr::select(-zipcode)
test_df  <- test_df  %>% dplyr::select(-zipcode)
```

## 1.2 Helper functions

```{r helper-functions}
# Evaluate predictions on the original price scale.
price_metrics <- function(pred_log, actual_price) {
  pred_price <- exp(as.numeric(pred_log))
  rmse <- sqrt(mean((actual_price - pred_price)^2))
  r2 <- 1 - sum((actual_price - pred_price)^2) /
    sum((actual_price - mean(actual_price))^2)

  tibble(RMSE = rmse, R2 = r2)
}

# Nice printing wrapper for console output.
print_price_metrics <- function(model_name, pred_log, actual_price) {
  out <- price_metrics(pred_log, actual_price)
  cat(model_name, "\n")
  cat("RMSE:", round(out$RMSE, 3), "\n")
  cat("R^2:", round(out$R2, 6), "\n\n")
  invisible(out)
}

# Extract the original term names selected by regsubsets.
# This supports the all-or-nothing treatment of factors.
extract_selected_terms <- function(selected_dummies, data) {
  mm <- model.matrix(log_price ~ . - price, data = data)
  mm_terms <- attr(terms(log_price ~ . - price, data = data), "term.labels")
  mm_assign <- attr(mm, "assign")[-1]
  mm_names <- colnames(mm)[-1]

  dummy_map <- tibble(
    dummy = mm_names,
    term_index = mm_assign,
    term = mm_terms[mm_assign]
  )

  unique(dummy_map$term[dummy_map$dummy %in% selected_dummies])
}

# Helper to display the largest coefficients.
extract_top_coefs <- function(coef_object, top_n = 12) {
  tibble(
    term = rownames(as.matrix(coef_object)),
    estimate = as.numeric(coef_object)
  ) %>%
    filter(term != "(Intercept)") %>%
    arrange(desc(abs(estimate))) %>%
    slice_head(n = top_n)
}
```

# 2. Best subset selection

Best subset selection performs an exhaustive search over models of different sizes. The rubric specifically asks for **Adjusted** $R^2$, **Mallows** $C_p$, and **BIC**, so those metrics are computed and plotted below.

```{r best-subset-fit}
# Number of candidate predictors after dummy encoding (excluding intercept).
p <- ncol(model.matrix(log_price ~ . - price, data = train_df)) - 1

best_fit <- regsubsets(
  log_price ~ . - price,
  data = train_df,
  nvmax = p,
  method = "exhaustive"
)

best_summary <- summary(best_fit)

best_metrics <- tibble(
  size = 1:p,
  adj_r2 = best_summary$adjr2,
  cp = best_summary$cp,
  bic = best_summary$bic
)

best_adjr2 <- which.max(best_summary$adjr2)
best_cp    <- which.min(best_summary$cp)
best_bic   <- which.min(best_summary$bic)

kable(
  tibble(
    Criterion = c("Adjusted R^2", "Cp", "BIC"),
    Best_Model_Size = c(best_adjr2, best_cp, best_bic)
  ),
  digits = 4,
  caption = "Optimal model sizes from best subset selection"
)
```

```{r best-subset-metrics-output}
cat("Regsubsets metrics\n")
cat("Best model size by Adjusted R^2:", best_adjr2, "\n")
cat("Best model size by Cp:", best_cp, "\n")
cat("Best model size by BIC:", best_bic, "\n")
```

```{r best-subset-plots, fig.width=11, fig.height=3.8}
par(mfrow = c(1, 3))

plot(best_metrics$size, best_metrics$adj_r2,
     type = "b", xlab = "Number of predictors", ylab = "Adjusted R^2")
points(best_adjr2, best_metrics$adj_r2[best_adjr2], col = "red", pch = 19)

plot(best_metrics$size, best_metrics$cp,
     type = "b", xlab = "Number of predictors", ylab = "Cp")
points(best_cp, best_metrics$cp[best_cp], col = "red", pch = 19)

plot(best_metrics$size, best_metrics$bic,
     type = "b", xlab = "Number of predictors", ylab = "BIC")
points(best_bic, best_metrics$bic[best_bic], col = "red", pch = 19)

par(mfrow = c(1, 1))
```

## 2.1 Interpretation of Adjusted $R^2$, $C_p$, and BIC

-   **Adjusted** $R^2$ rewards model fit but penalizes unnecessary predictors, so higher values are preferred.
-   **Mallows** $C_p$ measures the bias-variance tradeoff, so lower values are preferred.
-   **BIC** penalizes model size more strongly than Adjusted $R^2$ or $C_p$, so lower values are preferred when the goal is a more parsimonious model.

Because BIC applies the strongest complexity penalty, the **BIC-selected model** is used as the final best-subset candidate for test-set evaluation.

## 2.2 Refit the BIC-selected subset

The raw `regsubsets()` output is based on dummy-coded columns. To make the final model interpretable, the selected dummy variables are mapped back to their original terms, and factor variables are included using an all-or-nothing rule.

```{r best-subset-selected-terms}
best_bic_coefs <- coef(best_fit, id = best_bic)
selected_dummies <- names(best_bic_coefs)[-1]
selected_terms <- extract_selected_terms(selected_dummies, train_df)

selected_dummies
selected_terms
```

```{r best-subset-refit}
formula_bss <- as.formula(
  paste("log_price ~", paste(selected_terms, collapse = " + "))
)

bss_model <- lm(formula_bss, data = train_df)
summary(bss_model)
```

```{r best-subset-test-metrics}
pred_log_bss <- predict(bss_model, newdata = test_df)
bss_test_metrics <- print_price_metrics(
  "Best subset model (BIC-selected) test metrics:",
  pred_log_bss,
  test_df$price
)
```

### Best subset interpretation

The best subset model is expected to keep predictors related to **size**, **quality**, **location**, and **amenities**, which are the dominant drivers of housing price in this dataset. In the presentation, the best BIC model was also the strongest held-out performer, with approximately **RMSE = 142,317** and **test \$R\^2 = 0.816**. Those results are consistent with a strong but still interpretable model.

# 3. Stepwise selection

Stepwise selection is less computationally expensive than exhaustive search because it adds and removes predictors sequentially rather than evaluating every possible combination.

```{r stepwise-fit}
full_model <- lm(log_price ~ . - price, data = train_df)
step_model <- step(full_model, direction = "both", trace = FALSE)

summary(step_model)
```

```{r stepwise-top-predictors}
step_coefs <- tidy(step_model) %>%
  filter(term != "(Intercept)") %>%
  arrange(p.value)

kable(
  step_coefs %>% slice_head(n = 12),
  digits = 4,
  caption = "Most statistically significant terms in the stepwise model"
)
```

```{r stepwise-test-metrics}
pred_log_step <- predict(step_model, newdata = test_df)
step_test_metrics <- print_price_metrics(
  "Stepwise model test metrics:",
  pred_log_step,
  test_df$price
)
```

### Stepwise interpretation and comparison with best subset

The stepwise model should select many of the same structural and location-based predictors chosen by best subset selection. In the current presentation, the stepwise model emphasizes `zipcode_log_median`, `sqft_living`, `lat`, `long`, `yr_built`, `waterfront`, and higher `view` categories, which is substantively consistent with the best subset story. Its reported held-out performance is also very similar to best subset, with approximately **RMSE = 142,518** and **test \$R\^2 = 0.8156**.

# 4. Regularization with 5-fold cross-validation

Because the response is continuous, the regularization analysis uses **ridge regression** and **lasso regression**. Both are fit with **5-fold cross-validation**, as required by the rubric.

```{r design-matrices}
x_train <- model.matrix(log_price ~ . - log_price - price, data = train_df)[, -1]
y_train <- train_df$log_price

x_test <- model.matrix(log_price ~ . - log_price - price, data = test_df)[, -1]
y_test <- test_df$log_price

dim(x_train)
dim(x_test)
```

## 4.1 Ridge regression

```{r ridge-cv, fig.width=7, fig.height=5}
cv_ridge <- cv.glmnet(
  x_train,
  y_train,
  alpha = 0,
  nfolds = 5
)

plot(cv_ridge)
title("Ridge regression: 5-fold CV error curve")

ridge_lambda_min <- cv_ridge$lambda.min
ridge_lambda_1se <- cv_ridge$lambda.1se

ridge_lambda_min
ridge_lambda_1se
```

```{r ridge-coefficients}
ridge_coef_min <- coef(cv_ridge, s = "lambda.min")
ridge_coef_1se <- coef(cv_ridge, s = "lambda.1se")

kable(
  extract_top_coefs(ridge_coef_min, top_n = 12),
  digits = 5,
  caption = "Largest ridge coefficients at lambda.min"
)
```

```{r ridge-test-metrics}
pred_log_ridge_min <- predict(cv_ridge, s = "lambda.min", newx = x_test)
pred_log_ridge_1se <- predict(cv_ridge, s = "lambda.1se", newx = x_test)

ridge_test_min <- print_price_metrics(
  "Ridge model test metrics (lambda.min):",
  pred_log_ridge_min,
  test_df$price
)

ridge_test_1se <- print_price_metrics(
  "Ridge model test metrics (lambda.1se):",
  pred_log_ridge_1se,
  test_df$price
)
```

### Ridge interpretation

`lambda.min` gives the value of $\lambda$ with the lowest cross-validated error, while `lambda.1se` gives the largest $\lambda$ within one standard error of the minimum. The latter is often preferred when a slightly simpler and more stable model is desired. In the current presentation, ridge at `lambda.min` achieves approximately **RMSE = 142,363** and **test \$R\^2 = 0.816**, very close to best subset.

## 4.2 LASSO regression

```{r lasso-cv, fig.width=7, fig.height=5}
cv_lasso <- cv.glmnet(
  x_train,
  y_train,
  alpha = 1,
  nfolds = 5
)

plot(cv_lasso)
title("LASSO regression: 5-fold CV error curve")

lasso_lambda_min <- cv_lasso$lambda.min
lasso_lambda_1se <- cv_lasso$lambda.1se

lasso_lambda_min
lasso_lambda_1se
```

```{r lasso-coefficients}
lasso_coef_min <- coef(cv_lasso, s = "lambda.min")
lasso_coef_1se <- coef(cv_lasso, s = "lambda.1se")

zero_predictors_min <- rownames(lasso_coef_min)[as.numeric(lasso_coef_min) == 0]
nonzero_predictors_min <- rownames(lasso_coef_min)[as.numeric(lasso_coef_min) != 0]

kable(
  extract_top_coefs(lasso_coef_min, top_n = 12),
  digits = 5,
  caption = "Largest LASSO coefficients at lambda.min"
)

zero_predictors_min
```

```{r lasso-test-metrics}
pred_log_lasso_min <- predict(cv_lasso, s = "lambda.min", newx = x_test)
pred_log_lasso_1se <- predict(cv_lasso, s = "lambda.1se", newx = x_test)

lasso_test_min <- print_price_metrics(
  "LASSO model test metrics (lambda.min):",
  pred_log_lasso_min,
  test_df$price
)

lasso_test_1se <- print_price_metrics(
  "LASSO model test metrics (lambda.1se):",
  pred_log_lasso_1se,
  test_df$price
)
```

### LASSO interpretation

Unlike ridge, lasso can shrink some coefficients exactly to zero, which makes it a built-in variable-selection method. In the current presentation, lasso at `lambda.min` sets **`condition3`** and **`grade8`** to zero and achieves approximately **RMSE = 142,517** and **test \$R\^2 = 0.8156**.

# 5. Model comparison and final justification

The rubric asks for a direct comparison across subset, stepwise, ridge, and lasso methods. The table below compares the main held-out models. For the regularized methods, the primary comparison uses **`lambda.min`** because that corresponds to the minimum cross-validated error.

```{r final-comparison-table}
comparison_table <- bind_rows(
  tibble(
    Method = "Best Subset (BIC)",
    Model_Size_or_Tuning = paste0("size = ", best_bic),
    RMSE = bss_test_metrics$RMSE,
    R2 = bss_test_metrics$R2
  ),
  tibble(
    Method = "Stepwise",
    Model_Size_or_Tuning = "both directions",
    RMSE = step_test_metrics$RMSE,
    R2 = step_test_metrics$R2
  ),
  tibble(
    Method = "Ridge",
    Model_Size_or_Tuning = paste0("lambda.min = ", signif(ridge_lambda_min, 5)),
    RMSE = ridge_test_min$RMSE,
    R2 = ridge_test_min$R2
  ),
  tibble(
    Method = "LASSO",
    Model_Size_or_Tuning = paste0("lambda.min = ", signif(lasso_lambda_min, 5)),
    RMSE = lasso_test_min$RMSE,
    R2 = lasso_test_min$R2
  )
) %>%
  arrange(RMSE)

kable(
  comparison_table,
  digits = 4,
  caption = "Final model comparison on the test set"
)
```

## 5.1 Final discussion

Overall, all four approaches are expected to perform very similarly on this dataset. That pattern is already visible in the current presentation, where the reported test-set performance is approximately: best subset **142,317 / 0.816**, ridge **142,363 / 0.816**, stepwise **142,518 / 0.8156**, and lasso **142,517 / 0.8156** for RMSE / test $R^2$, respectively.

The **BIC-selected best subset model** is the strongest final choice because it provides:

-   the **lowest held-out RMSE** in the current results;
-   a strong test $R^2$;
-   an interpretable set of predictors centered on **location**, **living area**, **quality**, **view**, and **waterfront status**.

At the same time, the tiny differences across methods are important to mention: ridge is almost identical in predictive performance, which suggests the overall signal in the data is stable and not overly dependent on one modeling strategy. This is a useful point for the presentation because it shows that the findings are not fragile.

## 5.2 Overfitting and the role of cross-validation

Subset and stepwise procedures can overfit if model complexity is chosen only from training fit. That is why the project uses **BIC** for a stronger complexity penalty in best subset selection and **5-fold cross-validation** for ridge and lasso. Cross-validation helps estimate out-of-sample error and therefore improves generalization. The similarity between train-selected models and test-set results supports the conclusion that the final models are not severely overfit.

# 6. Classification extension

This project is a **regression** project with a continuous outcome, so the classification extension in the rubric is **not applicable here**. If the group later adds a binary classification target (for example, above-vs-below median price), then confusion matrix, sensitivity, specificity, ROC curve, and AUC could be included in a separate extension analysis. For the current submission, the core required work is the regression comparison.

# 7. Final conclusion

Sale price in the King County housing dataset is driven primarily by **location**, **home size**, **quality**, **waterfront access**, and **view-related amenities**. Across best subset, stepwise, ridge, and lasso, these same themes repeatedly emerge.

The final recommended model is the **BIC-selected best subset model** because it achieved the strongest test performance while remaining interpretable. Ridge and lasso confirm that the signal is stable under shrinkage, and 5-fold cross-validation provides additional protection against overfitting. This gives a strong, rubric-aligned justification for the final model choice.
